QuickSched: Task-based parallelism with dependencies and conflicts

### 2014-07-29 paper/paper.tex
refactoring introduction, added design considerations.

% v2-acmsmall-sample.tex, dated March 6 2012
% This is a sample file for ACM small trim journals
%
% Compilation using 'acmsmall.cls' - version 1.3 (March 2012), Aptara Inc.
% (c) 2010 Association for Computing Machinery (ACM)
%
% Questions/Suggestions/Feedback should be addressed to => "acmtexsupport@aptaracorp.com".
% Users can also go through the FAQs available on the journal's submission webpage.
%
% Steps to compile: latex, bibtex, latex latex
%
% For tracking purposes => this is v1.3 - March 2012

\documentclass[prodmode,acmtopc]{acmsmall} % Aptara syntax

@@

% Metadata Information
\acmVolume{0}
\acmNumber{0}
\acmArticle{0}
\acmYear{2014}
\acmMonth{0}

% My own macros
\usepackage{epsfig}
%\usepackage{hyperref}
\usepackage{listings}
\lstset{%
    language=C,
    basicstyle=\small\tt,
    numbers=left,
    numberstyle=\tiny
    }
\newcommand{\oh}[1]
    {\mbox{$ {\mathcal O}( #1 ) $}}
\newcommand{\eqn}[1]
    {(\ref{eqn:#1})}
\newcommand{\fig}[1]
    {Figure~\ref{fig:#1}}
\newcommand{\tabl}[1]
    {Table~\ref{tab:#1}}
\newcommand{\figs}[2]
    {Figures~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\sect}[1]
    {Section~\ref{sec:#1}}
\newcommand{\cuda}
    {{\sc cuda}\xspace}
\newcommand{\namd}
    {{\sc namd}\xspace}
\newcommand{\acemd}
    {{\sc acemd}\xspace}
\newcommand{\fenzi}
    {{\sc fenzi}\xspace}
\newcommand{\mdcore}
    {{\tt mdcore}\xspace}
\newcommand{\swift}
    {{\sc swift}\xspace}
\newcommand{\bsf}[1]
    {\textbf{\textsf{#1}}}

@@

% Page heads
\markboth{P.~Gonnet}{QuickSched: Task-based parallelism with dependencies and conflicts}

% Title portion
\title{QuickSched: Task-based parallelism with dependencies and conflicts}
\author{PEDRO GONNET
\affil{Durham University}}
% NOTE! Affiliations placed here should be for the institution where the
%       BULK of the research was done. If the author has gone to a new
%       institution, before publication, the (above) affiliation should NOT be changed.
%       The authors 'current' address may be given in the "Author's addresses:" block (below).
%       So for example, Mr. Abdelzaher, the bulk of the research was done at UIUC, and he is
%       currently affiliated with NASA.

@@

\category{???}{Computing methodologies}{Shared memory algorithms \and Concurrent algorithms}

\terms{Algorithms, Design, Performance}

\keywords{Task-based parallelism}

\acmformat{Pedro Gonnet, 2014.
QuickSched: Task-based parallelism with dependencies and conflicts.}
% At a minimum you need to supply the author names, year and a title.
% IMPORTANT:
% Full first names whenever they are known, surname last, followed by a period.
% In the case of two authors, 'and' is placed between them.
% In the case of three or more authors, the serial comma is used, that is, all author names
% except the last one but including the penultimate author's name are followed by a comma,
% and then 'and' is placed before the final author's name.
% If only first and middle initials are known, then each initial
% is followed by a period and they are separated by a space.
% The remaining information (journal title, volume, article number, date, etc.) is 'auto-generated'.

\begin{bottomstuff}
This work was supported by a Durham University Seedcorn Grant.

Author's address: P. Gonnet, School of Engineering and Computing Sciences,
Durham University, South Road, Durham, DH1 3LE, United Kingdom.
\end{bottomstuff}

@@

One of the first implementations of a task-based parallel programming
systems is Cilk \cite{ref:Blumofe1995}, an extension to the C
programming language which allows function calls to be ``spawned''
as new tasks.
Dependencies are enforced by the {\tt sync} keyword, which
forces a thread to wait for all the tasks that it spawned
to complete.
Although simple to use, this implicit dependency management
limits the types of DAGs that can be represented, i.e.~for
the example in \fig{Tasks}, using such a spawning model
would create implicit dependencies between the lowest-level
tasks $C$, $E$, and $K$.\footnote{
The main thread spawns tasks $A$, $G$ and $J$, $A$ spawns $B$ and $D$,
$G$ spawns $F$, $H$, and then $I$, $B$ spawns $C$.
The main thread then has to {\tt sync} for $A$, $G$, and $J$,
and thus implicitly all their spawned tasks, before executing
$E$ and $K$.}

@@

In dependency-only systems, such conflicts can be modelled
with dependencies, which enforce a pre-determined arbitrary
ordering on conflicting tasks.
This unnecessary restriction on the order
in which tasks can be scheduled can severely limit the
parallelizability of a computation, especially in the presence
of multiple conflicts per task.
Both \citeN{ref:Ltaief2012} and \citeN{ref:Agullo2013} note
this problem in their respective implementations of the Fast Multipole
Method (FMM), in which forces computed in different tasks are
accumulated on a set of particles.

@@

In order to be of practical use, and to scale well with modern
multi-core shared-memory architectures, any task scheduler
implementing constraints must be:
\begin{itemize}
    \item {\em Correct}: All constraints, i.e.~dependencies and
        conflicts, must be correctly enforced,
    \item {\em Fast}: The overheads associated with task management
        should be as small as possible,
    \item {\em Memory/cache-efficient}: Tasks accessing similar
        sets of data should be preferentially executed on the
        same core to preserve memory/cache locality as far as possible, and
    \item {\em Parallel-efficient}: Tasks should be executed in an order
        that sufficient work is available for all computational
        threads at all times.
\end{itemize}
\noindent This paper presents QuickSched, a framework for task-based
parallel programming with constraints, which aims to achieve
all of these goals.
Section~2 describes the underlying algorithms and data structures,
and Section~3 describes their specific implementation in QuickSched.
Section~4 presents two test-cases:
\begin{enumerate}
    \item The tiled QR
    decomposition described in \cite{ref:Buttari2009} and for
    which the QUARK scheduler was originally developed, and
    \item A task-based Barnes-Hut tree-code to compute the
    gravitational N-body problem similar to the FMM codes
    of \citeN{ref:Ltaief2012} and \citeN{ref:Agullo2013},
\end{enumerate}
These real-world examples show how QuickSched can be used in practice,
and can be used to assess its efficiency.
Section~5 concludes with some general observations and future work
directions.

@@

@@

Finally, {\tt cost} and {\tt weight} are measures
for the relative computational cost of this task, and the
relative cost of the critical path following the
dependencies of this task, respectively.
The task weights can be computed by first sorting
the tasks topologically according to their dependencies, i.e.
as per \citeN{ref:Kahn1962}:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
int top[N], j, k, sorted = 0, maxweight;
set the task waits as described above.
for ( k = 0 ; k < N ; k++ )
  if ( tasks[k].wait == 0 )
    top[ sorted++ ] = k;
for ( k = 0 ; k < sorted ; k++ )
  for ( j = 0 ; j < tasks[ top[k] ].nr_unlocks ; j++ )
    if ( ( tasks[ top[k] ].unlock[j]->wait -= 1 ) == 0 )
      top[ sorted++ ] = k;
if ( k < N )
  error( "Circular dependencies detected." );
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the array {\tt top} contains the task indices
in topological order. 
The test in line~10 is a convenient check if the tasks and their
dependencies actually do form an acyclic graph.
The weights themselves are then computed as follows
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
for ( k = N-1 ; k >= 0 ; k-- ) {
  for ( w = 0 , j = 0 ; j < tasks[ top[k] ].nr_unlocks ; j++ ) {
    tasks[ top[k] ].unlock[j]->wait += 1;
    w = max( tasks[ top[k] ].unlock[j]->weight , w );
    }
  tasks[ top[k] ].weight = tasks[ top[k] ].cost + w;
  }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the tasks are traversed in reverse
topological order, computing the recursive weight as the sum of the
task cost and the maximum weight of the tasks it unlocks (line~6),
and recomputing the task waits at the same time (line~3).

@@

\section{User Interface}

@@

The first test is the tiled QR decomposition originally
described in \citeN{ref:Buttari2009}, which has been used as a benchmark
by other authors \cite{ref:Agullo2009b,ref:Badia2009,ref:Bosilca2012}.
This example only requires dependencies and is presented 
as a point of comparison to existing task-based parallel
programming infrastructures.

The second example is a Barnes-Hut tree-code, a problem
similar to the Fast Multipole Method described in both
\citeN{ref:Ltaief2012} and \citeN{ref:Agullo2013}.
This example shows how conflicts, modeled
via hierarchical resources, can be useful.

@@

\citeN{ref:Buttari2009} introduced the concept of using task-based
parallelism for tile-based algorithms in numerical linear algebra,
presenting parallel codes for the Cholesky, LU, and QR
decompositions.
These algorithms are now part of the PLASMA and MAGMA
libraries for parallel linear algebra \cite{ref:Agullo2009}.
The former uses the QUARK task scheduler, which was originally
designed for it, while the latter currently uses the StarPU
task scheduler.

@@

\begin{center}
    \begin{tabular}{llll}
        Task & where & depends on task(s) & locks tile(s) \\
        \hline
        \epsfig{file=figures/TaskRed.pdf,height=9pt} DGEQRF & $i=j=k$ & $(i,j,k-1)$ & $(i,j)$ \\
        \epsfig{file=figures/TaskGreen.pdf,height=9pt} DLARFT & $i=k$, $j>k$ & $(i,j,k-1)$, $(k,k,k)$ & $(i,j)$ \\
        \epsfig{file=figures/TaskBlue.pdf,height=9pt} DTSQRF & $i>k$, $j=k$ & $(i,j,k-1)$, $(i-1,j,k)$ & $(i,j)$, $(j,j)$ \\
        \epsfig{file=figures/TaskYellow.pdf,height=9pt} DSSRFT & $i>k$, $j>k$ & $(i,j,k-1)$, $(k,j,k)$, $(i,k,k)$ & $(i,j)$ \\
        \hline
    \end{tabular}
\end{center}

@@

\begin{figure}
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
enum { tDGEQRF , tDLARFT , tDTSQRF , tDSSRFT };
void make_tasks ( struct qsched *s , int m , int n ) {
  int i, j, k, data[3];
  qsched_task_t tid[ m*n ], tid_new;
  qsched_res_t rid[ m*n ];
  for ( k = 0 ; k < m*n ; k++ ) {
    tid[k] = qsched_task_none;
    rid[k] = qsched_addres( s , qsched_res_none );
    }
  for ( k = 0 ; k < m && k < n ; k++ ) {
    /* DGEQRF task at (k,k). */
    data[0] = k; data[1] = k; data[2] = k;
    tid_new = qsched_addtask( s , tDGEQRF , qsched_flags_none ,
      data , sizeof(int)*3 , 2 );
    qsched_addlock( s , tid_new , rid[ k*m + k ] )
    if ( tid[ k*m + k ] != qsched_task_none )
      qsched_addunlock( s , tid[ k*m + k ] , tid_new );
    tid[ k*m + k ] = tid_new;
    for ( j = k+1 ; j < n ; j++ ) {
      /* DLARFT task at (k,j). */
      data[0] = k; data[1] = j; data[2] = k;
      tid_new = qsched_addtask( s , tDLARFT , qsched_flags_none ,
        data , sizeof(int)*3 , 3 );
      qsched_addlock( s , tid_new , rid[ j*m + k ] )
      qsched_addunlock( s , tid[ k*m + k ] , tid_new );
      if ( tid[ j*m + k ] != qsched_task_none )
        qsched_addunlock( s , tid[ j*m + k ] , tid_new );
      tid[ j*m + k ] = tid_new;
      }
    for ( i = k+1 ; i < m ; i++ ) {
      /* DTSQRF task at (i,k). */
      data[0] = i; data[1] = k; data[2] = k;
      tid_new = qsched_addtask( s , tDTSQRF , qsched_flags_none ,
        data , sizeof(int)*3 , 3 );
      qsched_addlock( s , tid_new , rid[ k*m + i ] )
      qsched_addlock( s , tid_new , rid[ k*m + k ] )
      qsched_addunlock( s , tid[ k*m + k ] , tid_new );
      if ( tid[ k*m + i ] != qsched_task_none )
        qsched_addunlock( s , tid[ k*m + i ] , tid_new );
      tid[ k*m + i ] = tid_new;
      for ( j = k+1 ; j < n ; j++ ) {
        /* DSSRFT task at (i,j). */
        data[0] = i; data[1] = j; data[2] = k;
        tid_new = qsched_addtask( s , tDSSRFT , qsched_flags_none ,
          data , sizeof(int)*3 , 5 );
        qsched_addlock( s , tid_new , rid[ j*m + i ] )
        qsched_addunlock( s , tid[ j*m + k ] , tid_new );
        qsched_addunlock( s , tid[ k*m + i ] , tid_new );
        if ( tid[ j*m + i ] != qsched_task_none )
          qsched_addunlock( s , tid[ j*m + i ] , tid_new );
        tid[ j*m + i ] = tid_new;
        }
      }
    }
  }
    \end{lstlisting}
\end{minipage}\end{center}
\caption{Example code to generate the tasks for the tiled
    QR decomposition.}
\label{fig:CodeQR}
\end{figure}

The following loops mirror the task generation described in
Algorithm~2 of \citeN{ref:Buttari2009}.
For each level {\tt k} (line~10), a DGEQRF task is created
for tile $(k,k)$ (lines~13--14).
A lock is added for the newly created task on the
resource associated with the $(k,k)$th tile (line~15).
If a task exists at that position at the previous level
(line~16), a dependency is added between the old task and
the new (line~17), and the new task is stored in {\tt tid}
(line~18).
The remaining tasks are generated in the same way, with
their respective locks and dependencies.

@@

\begin{figure}
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
enum { tSELF , tPAIR_PP , tPAIR_PC , tCOM };
void make_tasks ( struct qsched *s , struct cell *ci , struct cell *cj ) {
  int j, k;
  qsched_task_t tid;
  struct cell *data[2];
  if ( cj == NULL ) {
    if ( ci->split && ci->count > n_task )
      for ( j = 0 ; j < 8 ; j++ ) {
        make_tasks( s , ci->progeny[j] , NULL );
        for ( k = j+1 ; k < 8 ; k++ )
          make_tasks( s , ci->progeny[j] , ci->progeny[k] );
        }
    else {
      tid = qsched_addtask( s , tSELF , qsched_flags_none ,
        &ci , sizeof(struct cell *) , ci->count*ci->count );
      qsched_addlock( s , tid , ci->res );
      if ( ci->split )
        qsched_addunlock( s , ci->com , tid );
      }
    }
  else if ( ci and cj are well separated ) {
      data[0] = ci; data[1] = cj;
      tid = qsched_addtask( s , tPAIR_PC , qsched_flags_none ,
        data , sizeof(struct cell *)*2 , ci->count );
      qsched_addlock( s , tid , ci->res );
      qsched_addunlock( s , cj->com , tid );
      data[0] = cj; data[1] = ci;
      tid = qsched_addtask( s , tPAIR_PC , qsched_flags_none ,
        data , sizeof(struct cell *)*2 , cj->count );
      qsched_addlock( s , tid , cj->res );
      qsched_addunlock( s , ci->com , tid );
      }
  else if ( ci->split && cj->split && 
     ci->count*cj->count > n_task*n_task )
    for ( j = 0 ; j < 8 ; j++ )
      for ( k = 0 ; k < 8 ; k++ )
        make_tasks( s , ci->progeny[j] , cj->progeny[k] );
  else {
    data[0] = ci; data[1] = cj;
    tid = qsched_addtask( s , tPAIR_PP , qsched_flags_none ,
      data , sizeof(struct cell *)*2 , ci->count*cj->count );
    qsched_addlock( s , tid , ci->res );
    qsched_addlock( s , tid , cj->res );
    if ( ci->split && cj->split ) {
      qsched_addunlock( s , ci->com , tid );
      qsched_addunlock( s , cj->com , tid );
      }
    }
  }
    \end{lstlisting}
\end{minipage}\end{center}
    \caption{C-like pseudo-code for recursive task creation
        for the Barnes-Hut tree-code.}
    \label{fig:MakeTasks}
\end{figure}

@@

\begin{figure}
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void comp_com ( struct cell *c ) {
  int j, k;
  c->com[0] = 0.0; c->com[1] = 0.0; c->com[2] = 0.0;
  c->mas = 0.0;
  if ( c->split )
    for ( k = 0 ; k < 8 ; k++ ) {
      struct cell *cp = c->progeny[k]
      for ( j = 0 ; j < 3 ; j++ ) c->com[j] += cp->com[j]*cp->mass;
      c->mass += cp->mass;
      }
  else
    for ( k = 0 ; k < 8 ; k++ ) {
      struct part *p = &c->parts[k];
      for ( j = 0 ; j < 3 ; j++ ) c->com[j] += p->x[j]*p->mass;
      c->mass += p->mass;
      }
  c->com[0] /= c->mass; c->com[1] /= c->mass; c->com[2] /= c->mass; 
  }

void comp_self ( struct cell *c ) {
  int j, k;
  if ( c->split )
    for ( j = 0 ; j < 8 ; j++ ) {
      comp_self( c->progeny[j] );
      for ( k = j+1 ; k < 8 ; k++ )
        comp_pair( c->progeny[j] , c->progeny[k] );
      }
  else
    for ( j = 0 ; j < c->count ; j++ )
      for ( k = j+1 ; k < c->count ; k++ )
        interact c->parts[j] and c->parts[k].
  }

void comp_pair ( struct cell *ci , struct cell *cj ) {
  int j, k;
  if ( ci and cj well separated ) {
    comp_pair_pc( ci , cj );
    comp_pair_pc( cj , ci );
    }
  else if ( ci->split && cj->split )
    for ( j = 0 ; j < 8 ; j++ )
      for ( k = 0 ; k < 8 ; k++ )
        comp_pair( ci->progeny[j] , cj->progeny[k] );
  else
    for ( j = 0 ; j < ci->count ; j++ )
      for ( k = 0 ; k < cj->count ; k++ )
        interact ci->parts[j] and cj->parts[k].
  }

void comp_pair_cp ( struct cell *ci , struct cell *cj ) {
  int k;
  for ( k = 0 ; k < ci->count ; k++ )
    interact ci->parts[k] and cj center of mass.
  }
    \end{lstlisting}
\end{minipage}\end{center}
    \caption{Task functions for the Barnes-Hut tree-code.}
    \label{fig:BHTasks}
\end{figure}

This Barnes-Hut tree-code was used to approximate the gravitational
N-Body problem for 1\,000\,000 particles with random coordinates
in $[0,1]^3$.
The parameters $n_\mathsf{max}=100$ and $n_\mathsf{task}=5000$
were used to generate the tasks, and cell pairs were considered
well separated if not directly adjacent.
Using the above scheme generated 161\,613 tasks, of which
512 self-interaction tasks, 18\,532 particle-particle interaction
task, 105\,120 particle-cell interaction tasks, and 37\,449
center of mass tasks.
A total of 179\,632 dependencies were generated, along with
142\,696 locks on 37\,449 resources.

For these tests, OpenMP parallelism was used and resource
re-owning was switched off.
Resource ownership was attributed by dividing the global
{\tt parts} array by the number of queues and assigning each cell's
{\tt res} to the fraction of the {\tt parts} array to which
the first of its own {\tt parts} belong.
The interactions  computed 10 times for each number of
cores, and the average thereof taken for the scaling and
efficiency results in \fig{BHResults}.
The timings are for {\tt qsched\_run}, including the cost of
{\tt qsched\_start}, which does not run in parallel.
Setting up the scheduler, tasks, and resources took, in all
cases, an average of 51.3\,ms.

@@

% Acknowledgments
\begin{acks}
The author would like to thank Sam Townsend for implementing the
kernels in the task-based QR decomposition as part of his MSc Thesis
at Durham University.
The author also thanks Lydia Heck of the Institute for
Computational Cosmology at Durham University for providing access
to, and expertise on, the COSMA cluster used in the performance
evaluation.
\end{acks}

% Bibliography
\bibliographystyle{ACM-Reference-Format-Journals}
\bibliography{quicksched}
                             % Sample .bib file with references that match those in
                             % the 'Specifications Document (V1.5)' as well containing
                             % 'legacy' bibs and bibs with 'alternate codings'.
                             % Gerry Murray - March 2012

% History dates
\received{February 2007}{March 2009}{June 2009}

\end{document}
% End of v2-acmsmall-sample.tex (March 2012) - Gerry Murray, ACM

>>>>

\documentclass[preprint]{elsarticle}

@@

% My own macros
\usepackage{epsfig}
\usepackage{url}
\usepackage{listings}
\lstset{%
    language=C,
    basicstyle=\small\tt,
    numbers=left,
    numberstyle=\tiny
    }
\newcommand{\oh}[1]
    {\mbox{$ {\mathcal O}( #1 ) $}}
\newcommand{\eqn}[1]
    {(\ref{eqn:#1})}
\newcommand{\fig}[1]
    {Figure~\ref{fig:#1}}
\newcommand{\tabl}[1]
    {Table~\ref{tab:#1}}
\newcommand{\figs}[2]
    {Figures~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\sect}[1]
    {Section~\ref{sec:#1}}
\newcommand{\cuda}
    {{\sc cuda}\xspace}
\newcommand{\namd}
    {{\sc namd}\xspace}
\newcommand{\acemd}
    {{\sc acemd}\xspace}
\newcommand{\fenzi}
    {{\sc fenzi}\xspace}
\newcommand{\mdcore}
    {{\tt mdcore}\xspace}
\newcommand{\swift}
    {{\sc swift}\xspace}
\newcommand{\bsf}[1]
    {\textbf{\textsf{#1}}}

@@

% Title portion
\title{QuickSched: Task-based parallelism with dependencies and conflicts}
\author[ecs,ggl]{Pedro Gonnet\corref{cor}} \ead{pedro.gonnet@durham.ac.uk}
\author[ecs]{Aidan B. G. Chalk} \ead{aidan.chalk@durham.ac.uk}
\author[icc]{Matthieu Schaller} \ead{matthieu.schaller@durham.ac.uk}
\cortext[cor]{Corresponding author}
\address[ecs]{School of Engineering and Computing Sciences, Durham University, South Road, Durham DH1 3LE, United Kingdom}
\address[icc]{Institute for Computational Cosmology, Durham University, South Road, Durham DH1 3LE, United Kingdom}
\address[ggl]{Google Switzerland GmbH, Brandshenkestr. 110, 8002 Z\"urich, Switzerland}

@@

\begin{keyword}
    task-based parallelism \sep shared-memory parallelism \sep high performance computing \sep parallel computing
\end{keyword}

@@

One of the first implementations of a task-based parallel programming
systems is Cilk \cite{ref:Blumofe1995}, an extension to the C
programming language which allows function calls to be ``spawned''
as new tasks.
Dependencies are enforced by the {\tt sync} keyword, which
forces a thread to wait for all the tasks that it spawned
to complete.

@@

In dependency-only systems, such conflicts can be modelled
with dependencies, which enforce a pre-determined arbitrary
ordering on conflicting tasks.
This unnecessary restriction on the order
in which tasks can be scheduled can severely limit the
parallelizability of a computation, especially in the presence
of multiple conflicts per task.
Both \cite{ref:Ltaief2012} and \cite{ref:Agullo2013} note
this problem in their respective implementations of the Fast Multipole
Method (FMM), in which forces computed in different tasks are
accumulated on a set of particles.

@@

In order to be of practical use, and to scale well with modern
multi-core shared-memory architectures, any task scheduler
implementing constraints must be:
\begin{itemize}
    \item {\em Correct}: All constraints, i.e.~dependencies and
        conflicts, must be correctly enforced,
    \item {\em Fast}: The overheads associated with task management
        should be as small as possible,
    \item {\em Memory/cache-efficient}: Tasks accessing similar
        sets of data should be preferentially executed on the
        same core to preserve memory/cache locality as far as possible, and
    \item {\em Parallel-efficient}: Tasks should be executed in an order
        that sufficient work is available for all computational
        threads at all times.
\end{itemize}
\noindent This paper presents QuickSched, a framework for task-based
parallel programming with constraints, which aims to achieve
all of these goals.
Section~2 describes the main design considerations, Section~3 the
underlying algorithms and data structures, and
Section~4 their specific implementation in QuickSched.
Section~5 presents two test-cases:
\begin{enumerate}
    \item The tiled QR
    decomposition described in \cite{ref:Buttari2009} and for
    which the QUARK scheduler was originally developed, and
    \item A task-based Barnes-Hut tree-code to compute the
    gravitational N-body problem similar to the FMM codes
    of \cite{ref:Ltaief2012} and \cite{ref:Agullo2013},
\end{enumerate}
These real-world examples show how QuickSched can be used in practice,
and can be used to assess its efficiency.
Section~6 concludes with some general observations and future work
directions.

@@

\section{Design Considerations}

\begin{itemize}
  \item Two flavours of dependency generation: Through spawning as in Cilk,
    or automatically as per QUARK, OmpSS, etc...
  \item Limitation of spawning with complex task hierarchies.
  \item Limitation of dependency guessing is sometimes over-protective dependencies.
  \item We let the user define dependencies explictly.
  \item Con: Some effort needed to get dependencies right,
  \item Pro: All dependencies known at the start of a computation, better
    scheduling decisions can be made (see QR).
  \item A word on task granularity...
\end{itemize}

From a programmer's perspective, there are two main paradigms for generating
task dependencies:
\begin{itemize}
  \item Implicitly via spawning and waiting, e.g.~as in Cilk, or
  \item Automatic extraction from data dependencies.
\end{itemize}

The first scheme, spawning and waiting, is arguably the simplest to
use.
For simple depedency structures in which each task depends on only a
single task, i.e.~if the task DAG is a tree, each task {\em spawns}, or
creates, its dependent tasks after its completion.
Hence for the tasks $A$--$E$ in \fig{Tasks}, task $A$ would spawn
tasks $B$ and $D$, task $B$ would spawn task $C$, and task $D$ would
spawn task $E$.
If a task has more than one dependency, e.g. tasks $D$--$F$ in \fig{Tasks},
then the task generation order is reversed: Task $E$ is executed first,
and first spawns tasks $D$ and $F$, and waits for both their completion
before doing its own computations.

Although simple to use, this implicit dependency management
limits the types of DAGs that can be represented, i.e.~for
all the tasks in \fig{Tasks}, using such a spawning and waiting model
would create implicit dependencies between the lowest-level
tasks $C$, $E$, and $K$.
The main thread would spawn tasks $A$, $G$ and $J$, $A$ spawns $B$ and $D$,
$G$ spawns $F$, $H$, and then $I$, $B$ spawns $C$.
The main thread then has to wait for $A$, $G$, and $J$,
and thus implicitly all their spawned tasks, before executing
$E$ and $K$.

Automatic dependency extraction works by enforcing dependencies
between tasks that access the same data.
These data dependencies are provided explicitly by the programmer, e.g.~by
describing which parameters to a task are input, output, and input/output.
The dependencies are enforced in the order in which the tasks are 
created.
This approach usually relies on compiler extensions, e.g.~ {\tt pragma}s
in C/C++, or a system of runtime wrappers, to describe the task parameters
and their intent.

This approach allows programmers to specify rich dependency hierarchies
with very little effort, they still only allow for one type of relationship,
i.e.~dependencies, and lack the ability to deal with conflicts as
described in the previous section.

Here, QuickSched takes a drastically different approach.
The programmer creates the task DAG explicitly before execution.

@@

Finally, {\tt cost} and {\tt weight} are measures
for the relative computational cost of this task, and the
relative cost of the critical path following the
dependencies of this task, respectively.
The task weights can be computed by first sorting
the tasks topologically according to their dependencies, i.e.
as per \cite{ref:Kahn1962}:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
int top[N], j, k, sorted = 0, maxweight;
set the task waits as described above.
for ( k = 0 ; k < N ; k++ )
  if ( tasks[k].wait == 0 )
    top[ sorted++ ] = k;
for ( k = 0 ; k < sorted ; k++ )
  for ( j = 0 ; j < tasks[ top[k] ].nr_unlocks ; j++ )
    if ( ( tasks[ top[k] ].unlock[j]->wait -= 1 ) == 0 )
      top[ sorted++ ] = k;
if ( k < N )
  error( "Circular dependencies detected." );
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the array {\tt top} contains the task indices
in topological order. 
The test in line~10 is a convenient check if the tasks and their
dependencies actually do form an acyclic graph.
The weights themselves are then computed as follows
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
for ( k = N-1 ; k >= 0 ; k-- ) {
  for ( w = 0 , j = 0 ; j < tasks[ top[k] ].nr_unlocks ; j++ ) {
    tasks[ top[k] ].unlock[j]->wait += 1;
    w = max( tasks[ top[k] ].unlock[j]->weight , w );
    }
  tasks[ top[k] ].weight = tasks[ top[k] ].cost + w;
  }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the tasks are traversed in reverse
topological order, computing the recursive weight as the sum of the
task cost and the maximum weight of the tasks it unlocks (line~6),
and recomputing the task waits at the same time (line~3).

@@

@@

The first test is the tiled QR decomposition originally
described in \cite{ref:Buttari2009}, which has been used as a benchmark
by other authors \cite{ref:Agullo2009b,ref:Badia2009,ref:Bosilca2012}.
This example only requires dependencies and is presented 
as a point of comparison to existing task-based parallel
programming infrastructures.

The second example is a Barnes-Hut tree-code, a problem
similar to the Fast Multipole Method described in both
\cite{ref:Ltaief2012} and \cite{ref:Agullo2013}.
This example shows how conflicts, modeled
via hierarchical resources, can be useful.

@@

\cite{ref:Buttari2009} introduced the concept of using task-based
parallelism for tile-based algorithms in numerical linear algebra,
presenting parallel codes for the Cholesky, LU, and QR
decompositions.
These algorithms are now part of the PLASMA and MAGMA
libraries for parallel linear algebra \cite{ref:Agullo2009}.
The former uses the QUARK task scheduler, which was originally
designed for it, while the latter currently uses the StarPU
task scheduler.

@@

\begin{center}
    \begin{tabular}{llll}
        Task & where & depends on task(s) & locks tile(s) \\
        \hline
        \epsfig{file=figures/TaskRed.pdf,height=9pt} DGEQRF & $i=j=k$ & $(i,j,k-1)$ & $(i,j)$ \\
        \epsfig{file=figures/TaskGreen.pdf,height=9pt} DLARFT & $i=k$, $j>k$ & $(i,j,k-1)$, $(k,k,k)$ & $(i,j)$ \\
        \epsfig{file=figures/TaskBlue.pdf,height=9pt} DTSQRF & $i>k$, $j=k$ & $(i,j,k-1)$, $(i-1,j,k)$ & $(i,j)$, $(j,j)$ \\
        \epsfig{file=figures/TaskYellow.pdf,height=9pt} DSSRFT & $i>k$, $j>k$ & $(i,j,k-1)$, $(i-1,j,k)$, $(i,k,k)$ & $(i,j)$ \\
        \hline
    \end{tabular}
\end{center}

@@

@@

@@

This Barnes-Hut tree-code was used to approximate the gravitational
N-Body problem for 1\,000\,000 particles with random coordinates
in $[0,1]^3$.
The parameters $n_\mathsf{max}=100$ and $n_\mathsf{task}=5000$
were used to generate the tasks, and cell pairs were considered
well separated if not directly adjacent.
Using the above scheme generated 97\,553 tasks, of which
512 self-interaction tasks, ??? particle-particle interaction
task, ??? particle-cell interaction tasks, and 37\,449
center of mass tasks.
A total of 141\,840 dependencies were generated, along with
104\,392 locks on 37\,449 resources.

For these tests, OpenMP parallelism was used and resource
re-owning was switched off.
Resource ownership was attributed by dividing the global
{\tt parts} array by the number of queues and assigning each cell's
{\tt res} to the fraction of the {\tt parts} array to which
the first of its own {\tt parts} belong.
The interactions were computed 10 times for each number of
cores, and the average thereof taken for the scaling and
efficiency results in \fig{BHResults}.
The timings are for {\tt qsched\_run}, including the cost of
{\tt qsched\_start}, which does not run in parallel.
Setting up the scheduler, tasks, and resources took, in all
cases, an average of 51.3\,ms.

@@

% Acknowledgments
\section*{Acknowledgments}
The author would like to thank Sam Townsend for implementing the
kernels in the task-based QR decomposition as part of his MSc Thesis
at Durham University.
The author also thanks Lydia Heck of the Institute for
Computational Cosmology at Durham University for providing access
to, and expertise on, the COSMA cluster used in the performance
evaluation.
This work was supported by a Durham University Seedcorn Grant.

% Bibliography
\bibliographystyle{elsarticle-num}
\bibliography{quicksched}

% Appendices
\appendix
\section{User Interface}

\section{Tiled QR Implementation}
